---
title: "End-to-end geospatial object detection with geodetect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end geospatial object detection with geodetect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set TRUE when running with real data
)
```

## Why geodetect?

Most R geospatial packages stop at image classification or pixel-level
segmentation (like `geodl`). `geodetect` fills the next gap: **object
detection** — finding and counting discrete objects in satellite imagery
with geographic bounding boxes. No Python, no conda environments — just
pure R.

Typical use cases:

- Counting buildings in post-disaster aerial photography
- Detecting vehicles in parking lots from drone imagery  
- Mapping agricultural field boundaries from Sentinel-2 imagery
- Monitoring deforestation by detecting forest gaps

---

## 0. Install

```{r install}
# Install from GitHub
# install.packages("remotes")
# remotes::install_github("yourusername/geodetect")

library(geodetect)
library(terra)
library(sf)

# Check GPU availability (strongly recommended for training)
has_gpu()
```

---

## 1. Prepare your data

You need:

1. A multi-band geospatial raster (GeoTIFF, Sentinel-2 scene, aerial ortho, etc.)
2. A vector layer (GeoPackage, shapefile) of **bounding box annotations**
   around objects of interest, with a `class` column.

```{r data}
# Load your raster scene (e.g. RGB aerial image or Sentinel-2 composite)
scene <- terra::rast("my_scene.tif")
print(scene)  # Check: bands, CRS, extent

# Load annotations (polygons drawn around objects in QGIS, ArcGIS, etc.)
labels <- sf::st_read("annotations.gpkg")
print(table(labels$class))  # Check class distribution

# Define your class map (class name → integer ID, 1-indexed)
class_map <- c(building = 1L, vehicle = 2L, tree = 3L)
```

---

## 2. Chip the raster

Break the large scene into 512×512 pixel chips with 25% overlap.

```{r chip}
chips_df <- make_chips(
  rast          = scene,
  chip_size     = 512,
  overlap       = 0.25,
  out_dir       = "chips/",
  min_valid     = 0.5   # Skip chips with > 50% NoData
)

chips_summary(chips_df)
head(chips_df)
```

---

## 3. Assign labels to chips

Map bounding box annotations to per-chip pixel coordinates.

```{r labels}
labels_list <- assign_labels_to_chips(
  chips_df      = chips_df,
  labels_sf     = labels,
  chip_size     = 512,
  iou_threshold = 0.3,   # Annotation must overlap chip by at least 30% IoU
  class_map     = class_map
)

# How many chips have at least one annotation?
n_pos <- sum(sapply(labels_list, nrow) > 0)
cat(n_pos, "chips with annotations out of", nrow(chips_df), "\n")
```

---

## 4. Split into train / validation

```{r split}
splits <- split_chips(chips_df, labels_list, val_frac = 0.2, seed = 42)

train_chips  <- splits$train$chips_df
train_labels <- splits$train$labels_list
val_chips    <- splits$val$chips_df
val_labels   <- splits$val$labels_list

cat("Train chips:", nrow(train_chips), "\n")
cat("Val chips:  ", nrow(val_chips),   "\n")
```

---

## 5. Create torch Datasets

```{r datasets}
train_ds <- geo_detection_dataset(
  chips_df    = train_chips,
  labels_list = train_labels,
  chip_size   = 512,
  n_bands     = 3,
  augment     = TRUE,    # Random flips + brightness jitter for training
  normalize   = TRUE     # ImageNet normalization (matches pretrained backbone)
)

val_ds <- geo_detection_dataset(
  chips_df    = val_chips,
  labels_list = val_labels,
  chip_size   = 512,
  n_bands     = 3,
  augment     = FALSE,
  normalize   = TRUE
)

cat("Training samples:", length(train_ds), "\n")
cat("Validation samples:", length(val_ds), "\n")
```

---

## 6. Build and train the detector

```{r train}
# Build Faster R-CNN with ImageNet pretrained ResNet-50 backbone
model <- build_detector(
  num_classes         = length(class_map),  # 3 classes (+ background internally)
  pretrained_backbone = TRUE,
  freeze_backbone     = FALSE,
  anchor_sizes        = list(c(16L), c(32L), c(64L), c(128L), c(256L)),
  # Smaller anchors for aerial imagery where objects are small
  min_size            = 512L,
  max_size            = 1024L
)

# Train — automatically saves the best checkpoint
result <- train_detector(
  model       = model,
  train_ds    = train_ds,
  val_ds      = val_ds,
  epochs      = 30,
  batch_size  = 4,     # Reduce to 2 if GPU OOM
  lr          = 0.005,
  patience    = 5,
  class_map   = class_map,
  chip_size   = 512,
  n_bands     = 3,
  save_path   = "geodetect_buildings.pt"
)

# Plot training curve
plot(result$history$train_loss, type = "l", col = "steelblue",
     xlab = "Epoch", ylab = "Loss", main = "Training curve")
lines(result$history$val_loss, col = "coral")
legend("topright", legend = c("Train", "Val"),
       col = c("steelblue", "coral"), lty = 1)
```

---

## 7. Evaluate (mAP)

```{r evaluate}
metrics <- evaluate_detector(
  model        = result$model,
  dataset      = val_ds,
  score_thresh = 0.5,
  device       = "cpu"
)

cat("mAP@50:      ", round(metrics$map50,    3), "\n")
cat("mAP@50:95:   ", round(metrics$map50_95, 3), "\n")
```

---

## 8. Run inference on a new scene

```{r predict}
# Load a new scene (can be a different extent or image)
new_scene <- terra::rast("new_area.tif")

# Load saved checkpoint
ckpt <- load_detector("geodetect_buildings.pt", device = "cpu")

# Sliding-window detection
detections <- predict_raster(
  model        = ckpt$model,
  rast         = new_scene,
  chip_size    = ckpt$chip_size,
  n_bands      = ckpt$n_bands,
  class_map    = ckpt$class_map,
  overlap      = 0.25,
  score_thresh = 0.5,
  nms_thresh   = 0.5
)

print(detections)
# sf object: class_id, class_name, score, geometry (POLYGON in scene CRS)
```

---

## 9. Visualise and export

```{r visualise}
# Plot detections overlaid on the scene
plot_detections(
  rast       = new_scene,
  detections = detections,
  title      = "Detected objects"
)

# Export to GeoPackage for use in QGIS or ArcGIS
export_detections(detections, "detections.gpkg", overwrite = TRUE)

# Summary statistics
table(detections$class_name)
```

---

## Tips for good results

**Small training data (< 500 annotated objects):**
Set `freeze_backbone = TRUE` in `build_detector()` — only the detection
head trains, preventing overfitting on small datasets.

**Very small objects (< 20px):**
Reduce anchor sizes: `anchor_sizes = list(c(4L), c(8L), c(16L), c(32L), c(64L))`.

**Class imbalance:**
If one class has far fewer examples, use a higher `iou_threshold` (e.g. 0.5)
in `assign_labels_to_chips()` to be more strict about which chips count.

**Large images without GPU:**
Use `batch_size = 1` in `predict_raster()` and smaller `chip_size = 256`.

**CRS mismatch:**
Ensure your raster and annotation vector layer share the same CRS:
```r
labels <- sf::st_transform(labels, terra::crs(scene))
```
